---
title: Final assesment: MD analysis of Chignoline protein
Sources: Lindorff-Larsen, K., et al. (2011). "How Fast-Folding Proteins Fold." Science, 334(6055), 517-520.
author: Ivan Markin
Date: 04.05.2026
---
```{r}
#Load necessary libraries
library(dplyr)
library(tidyr)
library(ggplot2)
library(plotly)
```


#1:
##Creating dataset and quick overlook
```{r}
#Variables: Time(ns), RMSD(A), Rg(nm), SASA, HBonds, Temp_Run, Energy, Pressure, Region, Status
md_values <- data.frame(
  Time_ns = 1:30,
  RMSD = c(0.45, 0.52, 0.68, 1.10, 1.45, 1.88, 2.10, 2.05, 1.95, 2.12, 
           2.25, 2.30, 2.15, 1.90, 1.60, 1.40, 1.20, 1.05, 0.95, 0.88, 
           0.75, 0.65, 0.55, 0.48, 0.52, 0.58, 0.62, 0.65, 0.61, 0.59),
  Rg_nm = c(0.72, 0.73, 0.75, 0.82, 0.88, 0.95, 0.98, 0.97, 0.96, 0.99, 
            1.02, 1.05, 1.01, 0.94, 0.88, 0.82, 0.78, 0.75, 0.74, 0.73, 
            0.72, 0.71, 0.70, 0.70, 0.71, 0.72, 0.72, 0.73, 0.72, 0.71),
  SASA = c(1250, 1260, 1285, 1350, 1420, 1500, 1550, 1540, 1530, 1560, 
           1600, 1620, 1580, 1500, 1420, 1350, 1300, 1280, 1270, 1260, 
           1250, 1240, 1230, 1225, 1235, 1245, 1255, 1260, 1255, 1250),
  HBonds = c(8, 8, 7, 5, 4, 3, 2, 2, 3, 2, 1, 1, 2, 4, 5, 6, 7, 8, 8, 9, 
             9, 10, 10, 10, 9, 9, 8, 8, 8, 9),
  Temp_Run = rep(c("300K_Run1", "300K_Run2", "300K_Run3"), each = 10),
  Potential_Energy = c(-350, -348, -345, -330, -320, -310, -305, -306, -308, -304,
                       -300, -298, -302, -315, -325, -335, -342, -348, -350, -352,
                       -355, -358, -360, -362, -360, -358, -355, -352, -354, -356),
  Pressure_bar = rep(1.013, 30),
  Protein_Region = rep(c("Turn", "Beta-Hairpin", "Tail"), 10),
  Stability = c(rep("Folded", 3), rep("Unfolding", 10), rep("Refolding", 7), rep("Folded", 10))
)

#Save the dataset
write.csv(md_values, "chignolin_md_data.csv", row.names = FALSE)

#Load the dataset back
protein_data <- read.csv("chignolin_md_data.csv")

#Verify the data structure
head(protein_data)
```

#2
##tidyr package
```{r}
#Separate: Splitting Temp_Run into Temperature and Run number
#Combining Stability and Region for a specific tag

protein_tidied <- protein_data %>%
  separate(Temp_Run, into = c("Temperature", "Run_ID"), sep = "_") %>%
  unite("State_Tag", Stability, Protein_Region, sep = ":")

#pivot_longer: Putting RMSD and Rg into one column for easier comparison
protein_long <- protein_tidied %>%
  pivot_longer(cols = c(RMSD, Rg_nm), names_to = "Metric", values_to = "Value")

# Showing the tidied data
head(protein_tidied)
```

#3
##diplyr
```{r}
#Keep only data where the protein is in a 'Folded' state
#Mutate:Calculate a simple 'Energy_per_HBond' ratio
#Summarise:Find the average SASA for each Temperature group
summary_table <- protein_tidied %>%
  filter(grepl("Folded", State_Tag)) %>%
  mutate(Energy_Ratio = Potential_Energy / HBonds) %>%
  group_by(Temperature) %>%
  summarise(Avg_SASA = mean(SASA), Min_Energy = min(Potential_Energy))

print(summary_table)
```

#4
##Descriptive statistics
```{r}
#Mean of Solvent Accessible Surface Area (SASA)
avg_sasa <- mean(protein_data$SASA)
#Standard deviation of the RMSD
sd_rmsd <- sd(protein_data$RMSD)
#Range of Potential energy
energy_range <- range(protein_data$Potential_Energy)

cat("Mean SASA:", avg_sasa, "\nSD of RMSD:", sd_rmsd, "\nEnergy Range:", energy_range)
```

#5
##Functions,loops,if/else statements
```{r}
#One function that takes another function as an argument
#This function applies a mathematical operation to the RMSD column
apply_to_rmsd <- function(data_vec, operation) {
  return(operation(data_vec))
}
#Using 'max' as the function argument
max_drift <- apply_to_rmsd(protein_data$RMSD, max)
max_drift

#One loop and one conditional statement
#This loop checks each row to see if the RMSD indicates a high or low deviation
drift_category <- c()
for (i in 1:nrow(protein_data)) {
  if (protein_data$RMSD[i] > 1.5) {
    drift_category[i] <- "Significant deviation"
  } else {
    drift_category[i] <- "Minimal deviation"
  }
}
protein_data$Drift_Level <- drift_category
protein_data$Drift_Level

#One function utilizing lazy evaluation
#The warning_msg is only processed if check is TRUE
stability_check <- function(val, check = FALSE, warning_msg) {
  if (check) {
    print(warning_msg)
  }
  return(val < 1.0) # Returns TRUE if RMSD is low (stable)
}
#This runs without crashing because check is FALSE, so stop is never called
is_it_stable <- stability_check(protein_data$RMSD[1:5], check = FALSE, warning_msg = stop("ALERT!"))
is_it_stable
```

#6
##GGplot2 + plotly
```{r}
#Histogram of hydrogen bonds
ggplot(protein_data, aes(x = HBonds)) +
  geom_histogram(binwidth = 1, fill = "steelblue", color = "white") +
  labs(title = "Distribution of dydrogen bonds in chignolin protein", x = "Count of Hbonds", y = "Frequency")

#Plot of RMSD over time
ggplot(protein_data, aes(x = Time_ns, y = RMSD)) +
  geom_line(color = "red", size = 1) +
  geom_point() +
  theme_minimal() +
  labs(title = "RMSD trajectory", x = "Time (ns)", y = "RMSD (Angstroms)")

#Boxplot of energy by stability status
ggplot(protein_data, aes(x = Stability, y = Potential_Energy, fill = Stability)) +
  geom_boxplot() +
  theme_bw() +
  labs(title = "Potential energy distribution over folding state")

```

Explanation:
1.The hydrogen bonds are inversely related to the RMSD. As the protein unfolds and the RMSD increases, the number of hydrogen bonds drops significantly (reaching a minimum of 1).
2. The RMSD trajectory shows a clear unfolding event between 5ns and 15ns (where RMSD rises above 2.0 Å) followed by a successful refolding event where it returns to ~0.5 Å.
3. The boxplot shows the energetic landscape of the chignoline protein. It confirms that the folded state occupies the lowest energy well, while the unfolding phase corresponds to a significant increase in potential energy, representing the collapse of stabilizing structural contacts.

```{r}

sasa_over_gyration <- plot_ly(data = protein_data,
                       x = ~Rg_nm, 
                       y = ~SASA,
                       marker = list(size = 10, color = "red", symbol = "circle"),
                       type = "scatter", mode = "lines+markers") %>%
  layout(
    title = list(text = "Solvent Accessible Surface Area (SASA) change
          during  the unfolding process"),
    xaxis = list(title = "Gyration radius", gridcolor = "white"),
    yaxis = list(title = "SASA", gridcolor = "white"),
    
                       
    paper_bgcolor = "white"
  )
sasa_over_gyration
```
Explamation: The Solvent Accessible Surface Area (SASA) increases during the unfolding phase, which is expected as the hydrophobic core of the protein becomes exposed to the water solvent.


Conclusions drawn from chignoline folding analysis:

This project successfully applied R language to analyze a 30 ns Molecular Dynamics simulation of the chignoline protein by applying various data science tools, such  results were obtained:

Structural stability:The RMSD trajectory indicates a high structural flexibility. The protein begins in a native state, undergoes a clear unfolding event (peaking at ~2.3 Å), and demonstrates its fast-folding nature by returning to its original configuration by the end of the simulation.

Geometric correlation:Through direct plot_ly() visualization, a positive correlation was found between Radius of gyration (Rg) and SASA. This confirms that as the protein unfolds, its surface area exposes to the solvent increases linearly.

Chemical driving forces:The analysis shows that the number of Hydrogen bonds is inversely proportional to structural drift. The loss of these bonds is the primary trigger for the unfolding phase, while their reformation stabilizes the final folded state.

Thermodynamic favorability: The Energy boxplot reveals that the folded state corresponds to the lowest potential energy well. This confirms that the native structure is the most thermodynamically stable state, while the unfolding phase represents a higher energetically, transitional state.

